Fractional Types

### 2016-12-18 typespp.tex
edit 4.1 and move Figure 8

We are now ready to turn the groupoid constructions from the previous section into an extension of $\Pi$ with new type constructors and combinators for creating and manipulating syntactic counterparts to $\order{p}$ and $\iorder{p}$.

@@

%%%%%
\subsection{Non-Dependent Version}

@@

The syntactic category $\ft$ of fractional types introduces type expressions for
iteration groupoids, symmetry groupoids, and their products. The combinators
that relate these fractional types are in Fig.~\ref{pifrac:comb}. The bottom
group includes identity and sequential composition~$\circledcirc$ combinators
and ensures that the combinators can be applied anywhere inside a product
$\circledast$. There is also a combinator that relates two types $\order{c_1}$
and $\order{c_2}$ via $\order{\alpha}$ whenever $\alpha : c_1 \isotwo
c_2$.
Indeed if $c_1$ and $c_2$ are considered equivalent via a 2-combinator then we
also consider their iteration groupoids to be equivalent. The top group ensures
that $\circledast$ carries a symmetric monoidal structure with combinators
witnessing the unit, commutativity, and associativity properties. The last two
pairs combinators $\eta-/\epsilon-$ and $\eta+/\epsilon+$ are inspired from the
definition of compact closed categories~\cite{ccc} which are symmetric monoidal
categories in which every object has a dual. They are the ones that witness the
``fractional'' nature of symmetry groupoids as motivated in Sec.~\ref{sub:products}.

The last new syntactic category in $\Pi^/$ is that of values which deserves some
background discussion. When the types denote sets, it is evident what it means
to have a value of a given type: it is just an element of the set. When types
denote groupoids, it is less clear what it means to have a value, especially
when the total number of values, as reported by the groupoid cardinality, is a
proper fraction. In that case we the values up to an equivalence relation. The
idea is not uncommon: in the conventional $\lambda$-calculus, we list
$\lambda x.x$ and $\lambda y.y$ as separate values of type
$\tau \rightarrow \tau$ and then provide a separate equivalence relation
($\alpha$-equivalence) to express the fact that these two values are
indistinguishable. The treatment in our setting is similar: for iteration
groupoids $\order{c}$, every distinct iterate $c^k$ is a listed as a separate
value with the understanding that some iterates are
$\isotwo$-equivalent. Similarly for symmetry groupoids $\iorder{c}$, every
morphism is listed as a value $1/c^k$ with the same understanding that that some
of these morphisms are $\isotwo$-equivalent.

@@

%%%%%
\subsection{Dependent Pointed Types}

>>>>

We are now ready to turn the groupoid constructions from the previous section
into an extension of $\Pi$ with new type constructors and combinators for
creating and manipulating syntactic counterparts to $\order{p}$ and
$\iorder{p}$.  Just adding the ``obvious'' types and combinators does
not work; however, as the failures are quite informative, we nevertheless go
through this first attempt.  This will help illustrate why we need extra
machinery in Sec.~\ref{sec:deppointedtypes}.

@@

%%%%%
\subsection{First attempt}

@@

The syntactic category $\ft$ of fractional types introduces type expressions
for iteration groupoids, symmetry groupoids, and their products. The
combinators that relate these fractional types are in Fig.~\ref{pifrac:comb}.
The bottom group includes identity and sequential composition~$\circledcirc$
combinators and ensures that the combinators can be applied anywhere inside a
product $\circledast$. There is also a combinator that relates two types
$\order{c_1}$ and $\order{c_2}$ via $\order{\alpha}$ whenever $\alpha : c_1
\isotwo c_2$.  Indeed if $c_1$ and $c_2$ are considered equivalent via a
$2$-combinator then we also consider their iteration groupoids to be equivalent.
The top group ensures that $\circledast$ carries a symmetric monoidal structure
with combinators witnessing the unit, commutativity, and associativity
properties. The last two pairs of combinators $\eta\!- / \epsilon-$ and
$\eta\!+ / \epsilon+$ are inspired from the definition of compact closed
categories~\cite{ccc} which are symmetric monoidal categories in which every
object has a dual. They are the ones that witness the ``fractional'' nature of
symmetry groupoids as motivated in Sec.~\ref{sub:products}.

The last new syntactic category in $\Pi^/$ is that of values which deserves some
additional discussion. When types denote sets, values of a type are clear:
they are just the elements of the set.  In the case when types are groupoids,
this is less clear, especially when the groupoid cardinality is a proper
fraction.  What we get instead are equivalence classes of values.
The idea is not uncommon: in the conventional $\lambda$-calculus, we list
$\lambda x.x$ and $\lambda y.y$ as separate values of type
$\tau \rightarrow \tau$ and then provide a separate equivalence relation
($\alpha$-equivalence) to express the fact that these two values are
indistinguishable. The treatment in our setting is similar: for iteration
groupoids $\order{c}$, every distinct iterate $c^k$ is a listed as a separate
value with the understanding that some iterates are
$\isotwo$-equivalent. Similarly for symmetry groupoids $\iorder{c}$, every
morphism is listed as a value $1/c^k$ with the same understanding that that some
of these morphisms are $\isotwo$-equivalent.

@@

%%%%%
\subsection{Dependent Pointed Types}\label{sec:deppointedtypes}
